Regional Inequalities and economic growth; cross section and times series analysis

### 2023-11-01 MSB104_GR_1_Final_Assignment_research_article.qmd
mer text

---
title: "Regional Inequalities and economic growth; cross section and times series analysis"
author: "Anine Therese Karlsen & Mona Lisa Jones"
format: pdf
abstract: "This assignment aims to acquire, process, and analyze sub-national GDP and population data for a subset of European countries. Calculate GDP per capita and explore regional inequity using various descriptive statistics and visualizations.Furthermore examining the relationship between regional development and inequality, employing cross sectional estimation techniques for the year 2010."
editor: visual
bibliography: [Econometrics.bib]
output-file: Assignment1.qmd
page-layout: full
number-sections: true
papersize: a4
---

@@

Using time series data and cross sectional observations we investigate the GDP per capita and GINI trends in Portugal, France, Hungary, Slovenia and Denmark in the time period 2000 to 2020. The aims is to acquire, process, and analyze sub-national GDP and population data for a subset of European countries. Calculate GDP per capita and explore regional inequity using various descriptive statistics and visualizations

# Literature review

Regional growth and inequality have been important topics of research in Europe for many years. In recent decades, the European Union has implemented policies aimed at reducing regional disparities and promoting economic growth across its member states. Furthermore, despite these efforts, significant disparities in income and economic development persist across regions in Europe [@lessmann2017].

When it comes to economic growth, several studies refer to numerous influencing factors. @gennaioli2014 and @crescenzi2012 states that regional growth, much like national growth, is significantly influenced by geography and human capital. Geography impacts trade, resource availability, and susceptibility to natural events. Human capital, encompassing skills and knowledge, drives innovation and productivity. Furthermore, @gennaioli2014 mentions that wealthier nations like Denmark in our case, tend to experience faster regional convergence, suggesting national prosperity aids regional growth [@gennaioli2014]. Convergence implies that poorer regions grow faster than richer ones, gradually reducing disparities in wealth and income. Also referred to as beta convergence in @lessmann2017 paper.

In the context of Europe, and our selected countries we might expect to see a wide variety of regional economic performances, especially within Hungary and Slovakia and between the western European countries. Capital market regulations also play a fundamental role, with effective frameworks promoting quicker regional convergence [@gennaioli2014]. Taken this in to consideration we might expect a faster growth rate after the joining of the EU in 2003 and 2004. That said the opening of the free labor market might also influence the migration of labor in Slovakia and Hungary. Not to mention the economic crisis hitting shortly after our eastern European countries joined the union.

Whilst their recent EU membership might have posed certain challenges for Eastern European countries during the financial crisis, it also offered tools and resources that aided in their recovery in Slovakia and Hungary. As to the western Europeans such as Denmark, Portugal and France we expect to see a sharp decline in the economic growth. Likewise, with the recent covid19 pandemic in terms of financial aspects.

Regarding the effects of the economic crisis, another paper notes that it led to a significant increase in regional unemployment rates across Europe. Reporting some regions experiencing unemployment rates as high as 30% [@iammarino2019]. The authors also discuss how the crisis exposed the vulnerability of regions that were heavily reliant on a single industry or sector, and how sheltered economies were better able to weather the storm [@iammarino2019]. Another affect that surprised us was that the economic crisis led to increase inequality. The study "*the effects of financial crisis on regional inequality*" indicate any type of financial crisis results in higher income inequality [@nguyen2022].

@lessmann2017 regional inequality study address the importance of studying regional inequality. In particular, its potential consequences, such as political tensions that can undermine social and political stability [@lessmann2017]. The paper also discusses the relationship between regional inequality and personal income inequality and conflict risk. Furthermore, the paper provides us with insights into the inverted U-shaped relationship between regional inequality and the level of economic development in different country groups [@lessmann2017]. The inverted U-shape also referred to as Kuznets Curve suggests that as an economy develops, regional inequality initially rises, reaches a peak, and then starts to decline. @lessmann2017 paper also finds that there is an N-shaped relationship between regional inequality and economic development, which means that regional inequality increases again after the inverted U-shaped pattern has been completed [@lessmann2017].

Lastly, @iammarino2019 discusses regional inequalities in European countries. It presents evidence that inter-regional inequality has been increasing in the European Union since the turn of the millennium. The authors argue that this is due to the existence of several groups of regional economies in Europe that are structurally very different from one another [@iammarino2019]. Additionally, the authors argue that there are some countries in the EU that are more evenly developed than others, and that a map of under performance or over-performance means less in a high-income but evenly developed country, such as in Denmark in our case [@iammarino2019]. On the other hand, France has been a subject of discussion and policymaking for many years, regarding its pockets of extreme wealth. (finner ikke kilde)

@@

### **Outlier discussion**

>>>>

---
title: "Regional Inequalities and economic growth; cross section and times series analysis"
author: "Anine Therese Karlsen & Mona Lisa Jones"
format: pdf
abstract: "This assignment aims to acquire, process, and analyze sub-national GDP and population data for a subset of European countries. Firstly (assignment 1), Calculateing GDP per capita and exploreing regional inequity using various descriptive statistics and visualizations.Secondly (assignment 2), examining the relationship between regional development and inequality, employing cross sectional estimation techniques for the year 2010."
editor: visual
bibliography: [Regional_inequality_23.bib]
output-file: Assignment1.qmd
page-layout: full
number-sections: true
papersize: a4
---

@@

Using time series data and cross sectional observations we investigate the GDP per capita and GINI trends in Portugal, France, Hungary, Slovenia and Denmark in the time period 2000 to 2020. The aims is to acquire, process, and analyze sub-national GDP and population data for a subset of European countries.

The primary objective is to gain knowledge of econometric methods used for science in business through our given topic economic growth and inequality measures. Both GDP per capita and the Gini coefficient are vital indicators in the field of economics and are frequently used in a wide array of studies to analyse and interpret the economic conditions and social disparities of different regions or countries. Their analysis provides crucial insights for policymakers, economists, and researchers in shaping and evaluating the effectiveness of socio-economic policies.

The research question ***"What are the regional economic growth and inequality measurements from 2000 to 2020, in selected European countries and how does transport infrastructure affect the results?"*** is key in guiding our study. Setting a solid foundation for a comprehensive study that not only measure our given parameters, but also critically examines the role of transportation infrastructure play in these dyniamics. @lessmann2017 states that transportation costs play an important role in agglomeration and income, and reduced transportation costs as a determinant of regional inequality transportation costs have the potential to reduce regional inequalities.

The sructure of our study is methodically designed to facilitate this exploration. In part 1A and B we calculate GDP per capita and explore regional inequity using various descriptive statistics and visualizations such as panel data (***Assignment_1***).

In part 2A and B we use cross sectional method to examining the relationship between regional development and inequality, employing cross sectional estimation techniques for the year 2010 (***Assignment_2***).

# Literature Review: Regional Economic Growth, Inequality, and Transport Infrastructure (2000-2020)

Why some places grow and prosper is a fundamental question in social science research. This question motivated Adam Smith\'s work, *An Inquiry into the Nature and Causes of the Wealth of Nations,* (Smith [1863](https://link-springer-com.galanga.hvl.no/article/10.1007/s11187-014-9574-4#ref-CR32 "Smith, A. (1863). An inquiry into the nature and causes of the wealth of nations. Edinburgh: A. & C Black.")) and has been a major influence of the *new economic geography* [@feldman2014]*. Furthermore,* regional growth and inequality have been important topics of research in Europe for many years. In recent decades, the European Union has implemented policies aimed at reducing regional disparities and promoting economic growth across its member states. Furthermore, despite these efforts, significant disparities in income and economic development persist across regions in Europe [@lessmann2017].

**Economic Growth and Regional Disparities**

Regional growth in Europe exhibits a complex pattern, influenced significantly by factors such as geography, human capital, and economic policies. Lessmann and Seidel\'s research offers insights into these dynamics, highlighting the persistence of income disparities across regions despite overarching economic advancement [@lessmann2017]. Wealthier nations like Denmark tend to experience quicker regional convergence, a phenomenon supported by Gennaioli et al.\'s findings [@gennaioli2014]. However, this is not uniformly experienced across all European countries, with Eastern European nations such as Hungary and Slovakia showcasing different growth trajectories, particularly after their EU accession.

**Financial crisis, and EU membership for Slovakia and Hungary**

The 2008 financial crisis significantly disrupted regional growth and exacerbated income inequalities across Europe. Nguyen et al. highlight that financial turmoil typically leads to increased income disparity, a trend observed globally [@nguyen2022].

Focusing on Europe, we expect varied regional economic performances, especially when comparing Eastern European countries such as Hungary and Slovakia to their Western counterparts. Given the role of capital market regulations in promoting regional convergence, outlined by Gennaioli et al., we anticipate a potential acceleration in growth for Eastern European countries following their EU accession in 2003 and 2004 [@gennaioli2014]. However, the financial crisis and subsequent labor market shifts may have influenced these trends.

EU membership provided Slovakia and Hungary with resources for economic recovery, while Western European countries, including Denmark, Portugal, and France, likely experienced significant economic declines due to the financial crisis, a pattern possibly mirrored in the recent COVID-19 pandemic. These complexities highlight the importance of a detailed analysis to understand regional economic growth and inequality in challenging times.

**Regional inequality**

Another affect that surprised us was that the economic crisis led to increase inequality. The study "*the effects of financial crisis on regional inequality*" indicate any type of financial crisis results in higher income inequality [@nguyen2022].

@lessmann2017 regional inequality study address the importance of studying regional inequality. In particular, its potential consequences, such as political tensions that can undermine social and political stability [@lessmann2017]. The paper also discusses the relationship between regional inequality and personal income inequality and conflict risk.

**Determinants of regional inequality**

In Lessmann and Seidel\'s exploration of regional inequality, the intricacies of this phenomenon unfold, revealing a multifaceted landscape. The study delineates how economic development carves an N-shaped trajectory in regional disparities, illustrating an initial surge in inequality, followed by a decrease, and an eventual uptick again, paralleling the region\'s economic progress. Geographic characteristics stand out as pivotal, with the natural and topographical elements of a region significantly swaying income distributions [@lessmann2017].

Urbanization emerges as a crucial determinant, where bustling, populous urban centres usually enjoy elevated income levels, contrasting starkly with the often-stark disparities witnessed in more rural and secluded areas [@lessmann2017]. The regulatory landscape too plays a formidable role; robust market regulations, particularly within the realms of capital and labour, are highlighted as vital catalysts fostering regional convergence and mitigating the chasms of inequality [@lessmann2017]. Collectively, these determinants underscore the complexity of regional inequality, prompting a call for comprehensive and nuanced strategies to navigate and address its multifarious dimensions.

*Transport infrastructure*

Transport infrastructure plays a pivotal role in shaping regional economic landscapes. Improved connectivity fosters economic activities, reduces travel time, and enhances accessibility to markets and resources. Studies indicates that regions with robust transport infrastructure tend to attract more investments, generate employment, and ultimately contribute to regional prosperity.

**Expected Inverted U- and N shaped curves**

Furthermore, the paper provides us with insights into the inverted U-shaped relationship between regional inequality and the level of economic development in different country groups [@lessmann2017]. The inverted U-shape also referred to as Kuznets Curve suggests that as an economy develops, regional inequality initially rises, reaches a peak, and then starts to decline.

@lessmann2017 paper also finds that there is an N-shaped relationship between regional inequality and economic development, which means that regional inequality increases again after the inverted U-shaped pattern has been completed [@lessmann2017].

@iammarino2019 discusses regional inequalities in European countries. It presents evidence that inter-regional inequality has been increasing in the European Union since the turn of the millennium. The authors argue that this is due to the existence of several groups of regional economies in Europe that are structurally very different from one another [@iammarino2019]. Additionally, the authors argue that there are some countries in the EU that are more evenly developed than others, and that a map of under performance or over-performance means less in a high-income but evenly developed country, such as in Denmark in our case [@iammarino2019]. On the other hand, France has been a subject of discussion and policy making for many years, regarding its pockets of extreme wealth. (finner ikke kilde)

Lastly, It is evident that regional economic growth and inequality in Europe from 2000 to 2020 have been influenced by a constellation of factors, with transport infrastructure emerging as a significant determinant. Earlier studies provides us with a foundational understanding of these dynamics, while also highlighting the nuanced impacts of economic crises on regional disparities. The integration of transport infrastructure into this discourse adds an additional layer of complexity, underscoring its potential as a catalyst for regional development and equality.

@@

### **Outline discussion**

### 2023-11-01 MSB104_GR_1_Final_Assignment_research_article.tex
mer text

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

@@

\begin{document}
\maketitle
\begin{abstract}
This assignment aims to acquire, process, and analyze sub-national GDP
and population data for a subset of European countries. Calculate GDP
per capita and explore regional inequity using various descriptive
statistics and visualizations.Furthermore examining the relationship
between regional development and inequality, employing cross sectional
estimation techniques for the year 2010.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, interior hidden, enhanced]}{\end{tcolorbox}}\fi

@@

While national GDP and GDP per capita are vital indicators of a
country's aggregate economic health, they do not shed light on how
wealth or income is distributed among its residents. A high national GDP
can, paradoxically, coexist with pockets of regional deprivation
(\textbf{lessmann2017?}).

The truth of this statement becomes more evident when taking a closer
look at sub-national data. Regional wealth disparities are of prime
concern, especially when crafting policies for equitable growth
(\textbf{lessmann2017?}). A country's macro-level prosperity does not
automatically guarantee that all its regions partake equally in this
wealth. By studying smaller regions within a country, it is possible to
get a more nuanced narrative about the state of regional economic
disparities (\textbf{lessmann2017?}).

Using time series data and cross sectional observations we investigate
the GDP per capita and GINI trends in Portugal, France, Hungary,
Slovenia and Denmark in the time period 2000 to 2020. The aims is to
acquire, process, and analyze sub-national GDP and population data for a
subset of European countries. Calculate GDP per capita and explore
regional inequity using various descriptive statistics and
visualizations

\hypertarget{literature-review}{%
\section{Literature review}\label{literature-review}}

Regional growth and inequality have been important topics of research in
Europe for many years. In recent decades, the European Union has
implemented policies aimed at reducing regional disparities and
promoting economic growth across its member states. Furthermore, despite
these efforts, significant disparities in income and economic
development persist across regions in Europe (\textbf{lessmann2017?}).

When it comes to economic growth, several studies refer to numerous
influencing factors. (\textbf{gennaioli2014?}) and
(\textbf{crescenzi2012?}) states that regional growth, much like
national growth, is significantly influenced by geography and human
capital. Geography impacts trade, resource availability, and
susceptibility to natural events. Human capital, encompassing skills and
knowledge, drives innovation and productivity. Furthermore,
(\textbf{gennaioli2014?}) mentions that wealthier nations like Denmark
in our case, tend to experience faster regional convergence, suggesting
national prosperity aids regional growth (\textbf{gennaioli2014?}).
Convergence implies that poorer regions grow faster than richer ones,
gradually reducing disparities in wealth and income. Also referred to as
beta convergence in (\textbf{lessmann2017?}) paper.

In the context of Europe, and our selected countries we might expect to
see a wide variety of regional economic performances, especially within
Hungary and Slovakia and between the western European countries. Capital
market regulations also play a fundamental role, with effective
frameworks promoting quicker regional convergence
(\textbf{gennaioli2014?}). Taken this in to consideration we might
expect a faster growth rate after the joining of the EU in 2003 and
2004. That said the opening of the free labor market might also
influence the migration of labor in Slovakia and Hungary. Not to mention
the economic crisis hitting shortly after our eastern European countries
joined the union.

Whilst their recent EU membership might have posed certain challenges
for Eastern European countries during the financial crisis, it also
offered tools and resources that aided in their recovery in Slovakia and
Hungary. As to the western Europeans such as Denmark, Portugal and
France we expect to see a sharp decline in the economic growth.
Likewise, with the recent covid19 pandemic in terms of financial
aspects.

Regarding the effects of the economic crisis, another paper notes that
it led to a significant increase in regional unemployment rates across
Europe. Reporting some regions experiencing unemployment rates as high
as 30\% (\textbf{iammarino2019?}). The authors also discuss how the
crisis exposed the vulnerability of regions that were heavily reliant on
a single industry or sector, and how sheltered economies were better
able to weather the storm (\textbf{iammarino2019?}). Another affect that
surprised us was that the economic crisis led to increase inequality.
The study ``\emph{the effects of financial crisis on regional
inequality}'' indicate any type of financial crisis results in higher
income inequality (\textbf{nguyen2022?}).

(\textbf{lessmann2017?}) regional inequality study address the
importance of studying regional inequality. In particular, its potential
consequences, such as political tensions that can undermine social and
political stability (\textbf{lessmann2017?}). The paper also discusses
the relationship between regional inequality and personal income
inequality and conflict risk. Furthermore, the paper provides us with
insights into the inverted U-shaped relationship between regional
inequality and the level of economic development in different country
groups (\textbf{lessmann2017?}). The inverted U-shape also referred to
as Kuznets Curve suggests that as an economy develops, regional
inequality initially rises, reaches a peak, and then starts to decline.
(\textbf{lessmann2017?}) paper also finds that there is an N-shaped
relationship between regional inequality and economic development, which
means that regional inequality increases again after the inverted
U-shaped pattern has been completed (\textbf{lessmann2017?}).

Lastly, (\textbf{iammarino2019?}) discusses regional inequalities in
European countries. It presents evidence that inter-regional inequality
has been increasing in the European Union since the turn of the
millennium. The authors argue that this is due to the existence of
several groups of regional economies in Europe that are structurally
very different from one another (\textbf{iammarino2019?}). Additionally,
the authors argue that there are some countries in the EU that are more
evenly developed than others, and that a map of under performance or
over-performance means less in a high-income but evenly developed
country, such as in Denmark in our case (\textbf{iammarino2019?}). On
the other hand, France has been a subject of discussion and policymaking
for many years, regarding its pockets of extreme wealth. (finner ikke
kilde)

@@

In this part we will report and interpret different types of essential
descriptive statistics. Measuring regional income inequality is
challenging due to heterogeneity of regions (\textbf{lessmann2017?}).
The number of regions in our data set varies largely in size and
population. Since the focus of this paper is purely growth and changes
in inequities over time, the variations of size and population density
becomes a minor issue because the country-level territorial
heterogeneity is fixed (\textbf{lessmann2017?}).

@@

\hypertarget{outlier-discussion}{%
\subsubsection{\texorpdfstring{\textbf{Outlier
discussion}}{Outlier discussion}}\label{outlier-discussion}}

@@

\hypertarget{references}{%
\section{References}\label{references}}

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\begin{document}
\maketitle
\begin{abstract}
This assignment aims to acquire, process, and analyze sub-national GDP
and population data for a subset of European countries. Firstly
(assignment 1), Calculateing GDP per capita and exploreing regional
inequity using various descriptive statistics and
visualizations.Secondly (assignment 2), examining the relationship
between regional development and inequality, employing cross sectional
estimation techniques for the year 2010.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, breakable]}{\end{tcolorbox}}\fi

@@

While national GDP and GDP per capita are vital indicators of a
country's aggregate economic health, they do not shed light on how
wealth or income is distributed among its residents. A high national GDP
can, paradoxically, coexist with pockets of regional deprivation
(Lessmann and Seidel 2017).

The truth of this statement becomes more evident when taking a closer
look at sub-national data. Regional wealth disparities are of prime
concern, especially when crafting policies for equitable growth
(Lessmann and Seidel 2017). A country's macro-level prosperity does not
automatically guarantee that all its regions partake equally in this
wealth. By studying smaller regions within a country, it is possible to
get a more nuanced narrative about the state of regional economic
disparities (Lessmann and Seidel 2017).

Using time series data and cross sectional observations we investigate
the GDP per capita and GINI trends in Portugal, France, Hungary,
Slovenia and Denmark in the time period 2000 to 2020. The aims is to
acquire, process, and analyze sub-national GDP and population data for a
subset of European countries.

The primary objective is to gain knowledge of econometric methods used
for science in business through our given topic economic growth and
inequality measures. Both GDP per capita and the Gini coefficient are
vital indicators in the field of economics and are frequently used in a
wide array of studies to analyse and interpret the economic conditions
and social disparities of different regions or countries. Their analysis
provides crucial insights for policymakers, economists, and researchers
in shaping and evaluating the effectiveness of socio-economic policies.

The research question \textbf{\emph{``What are the regional economic
growth and inequality measurements from 2000 to 2020, in selected
European countries and how does transport infrastructure affect the
results?''}} is key in guiding our study. Setting a solid foundation for
a comprehensive study that not only measure our given parameters, but
also critically examines the role of transportation infrastructure play
in these dyniamics. Lessmann and Seidel (2017) states that
transportation costs play an important role in agglomeration and income,
and reduced transportation costs as a determinant of regional inequality
transportation costs have the potential to reduce regional inequalities.

The sructure of our study is methodically designed to facilitate this
exploration. In part 1A and B we calculate GDP per capita and explore
regional inequity using various descriptive statistics and
visualizations such as panel data (\textbf{\emph{Assignment\_1}}).

In part 2A and B we use cross sectional method to examining the
relationship between regional development and inequality, employing
cross sectional estimation techniques for the year 2010
(\textbf{\emph{Assignment\_2}}).

\hypertarget{literature-review-regional-economic-growth-inequality-and-transport-infrastructure-2000-2020}{%
\section{Literature Review: Regional Economic Growth, Inequality, and
Transport Infrastructure
(2000-2020)}\label{literature-review-regional-economic-growth-inequality-and-transport-infrastructure-2000-2020}}

Why some places grow and prosper is a fundamental question in social
science research. This question motivated Adam Smith\textquotesingle s
work, \emph{An Inquiry into the Nature and Causes of the Wealth of
Nations,} (Smith
\href{https://link-springer-com.galanga.hvl.no/article/10.1007/s11187-014-9574-4\#ref-CR32}{1863})
and has been a major influence of the \emph{new economic geography}
(Feldman 2014)\emph{. Furthermore,} regional growth and inequality have
been important topics of research in Europe for many years. In recent
decades, the European Union has implemented policies aimed at reducing
regional disparities and promoting economic growth across its member
states. Furthermore, despite these efforts, significant disparities in
income and economic development persist across regions in Europe
(Lessmann and Seidel 2017).

\textbf{Economic Growth and Regional Disparities}

Regional growth in Europe exhibits a complex pattern, influenced
significantly by factors such as geography, human capital, and economic
policies. Lessmann and Seidel\textquotesingle s research offers insights
into these dynamics, highlighting the persistence of income disparities
across regions despite overarching economic advancement (Lessmann and
Seidel 2017). Wealthier nations like Denmark tend to experience quicker
regional convergence, a phenomenon supported by Gennaioli et
al.\textquotesingle s findings (Gennaioli et al. 2014). However, this is
not uniformly experienced across all European countries, with Eastern
European nations such as Hungary and Slovakia showcasing different
growth trajectories, particularly after their EU accession.

\textbf{Financial crisis, and EU membership for Slovakia and Hungary}

The 2008 financial crisis significantly disrupted regional growth and
exacerbated income inequalities across Europe. Nguyen et al.~highlight
that financial turmoil typically leads to increased income disparity, a
trend observed globally (Nguyen 2022).

Focusing on Europe, we expect varied regional economic performances,
especially when comparing Eastern European countries such as Hungary and
Slovakia to their Western counterparts. Given the role of capital market
regulations in promoting regional convergence, outlined by Gennaioli et
al., we anticipate a potential acceleration in growth for Eastern
European countries following their EU accession in 2003 and 2004
(Gennaioli et al. 2014). However, the financial crisis and subsequent
labor market shifts may have influenced these trends.

EU membership provided Slovakia and Hungary with resources for economic
recovery, while Western European countries, including Denmark, Portugal,
and France, likely experienced significant economic declines due to the
financial crisis, a pattern possibly mirrored in the recent COVID-19
pandemic. These complexities highlight the importance of a detailed
analysis to understand regional economic growth and inequality in
challenging times.

\textbf{Regional inequality}

Another affect that surprised us was that the economic crisis led to
increase inequality. The study ``\emph{the effects of financial crisis
on regional inequality}'' indicate any type of financial crisis results
in higher income inequality (Nguyen 2022).

Lessmann and Seidel (2017) regional inequality study address the
importance of studying regional inequality. In particular, its potential
consequences, such as political tensions that can undermine social and
political stability (Lessmann and Seidel 2017). The paper also discusses
the relationship between regional inequality and personal income
inequality and conflict risk.

\textbf{Determinants of regional inequality}

In Lessmann and Seidel\textquotesingle s exploration of regional
inequality, the intricacies of this phenomenon unfold, revealing a
multifaceted landscape. The study delineates how economic development
carves an N-shaped trajectory in regional disparities, illustrating an
initial surge in inequality, followed by a decrease, and an eventual
uptick again, paralleling the region\textquotesingle s economic
progress. Geographic characteristics stand out as pivotal, with the
natural and topographical elements of a region significantly swaying
income distributions (Lessmann and Seidel 2017).

Urbanization emerges as a crucial determinant, where bustling, populous
urban centres usually enjoy elevated income levels, contrasting starkly
with the often-stark disparities witnessed in more rural and secluded
areas (Lessmann and Seidel 2017). The regulatory landscape too plays a
formidable role; robust market regulations, particularly within the
realms of capital and labour, are highlighted as vital catalysts
fostering regional convergence and mitigating the chasms of inequality
(Lessmann and Seidel 2017). Collectively, these determinants underscore
the complexity of regional inequality, prompting a call for
comprehensive and nuanced strategies to navigate and address its
multifarious dimensions.

\emph{Transport infrastructure}

Transport infrastructure plays a pivotal role in shaping regional
economic landscapes. Improved connectivity fosters economic activities,
reduces travel time, and enhances accessibility to markets and
resources. Studies indicates that regions with robust transport
infrastructure tend to attract more investments, generate employment,
and ultimately contribute to regional prosperity.

\textbf{Expected Inverted U- and N shaped curves}

Furthermore, the paper provides us with insights into the inverted
U-shaped relationship between regional inequality and the level of
economic development in different country groups (Lessmann and Seidel
2017). The inverted U-shape also referred to as Kuznets Curve suggests
that as an economy develops, regional inequality initially rises,
reaches a peak, and then starts to decline.

Lessmann and Seidel (2017) paper also finds that there is an N-shaped
relationship between regional inequality and economic development, which
means that regional inequality increases again after the inverted
U-shaped pattern has been completed (Lessmann and Seidel 2017).

(\textbf{iammarino2019?}) discusses regional inequalities in European
countries. It presents evidence that inter-regional inequality has been
increasing in the European Union since the turn of the millennium. The
authors argue that this is due to the existence of several groups of
regional economies in Europe that are structurally very different from
one another (\textbf{iammarino2019?}). Additionally, the authors argue
that there are some countries in the EU that are more evenly developed
than others, and that a map of under performance or over-performance
means less in a high-income but evenly developed country, such as in
Denmark in our case (\textbf{iammarino2019?}). On the other hand, France
has been a subject of discussion and policy making for many years,
regarding its pockets of extreme wealth. (finner ikke kilde)

Lastly, It is evident that regional economic growth and inequality in
Europe from 2000 to 2020 have been influenced by a constellation of
factors, with transport infrastructure emerging as a significant
determinant. Earlier studies provides us with a foundational
understanding of these dynamics, while also highlighting the nuanced
impacts of economic crises on regional disparities. The integration of
transport infrastructure into this discourse adds an additional layer of
complexity, underscoring its potential as a catalyst for regional
development and equality.

@@

In this part we will report and interpret different types of essential
descriptive statistics. Measuring regional income inequality is
challenging due to heterogeneity of regions (Lessmann and Seidel 2017).
The number of regions in our data set varies largely in size and
population. Since the focus of this paper is purely growth and changes
in inequities over time, the variations of size and population density
becomes a minor issue because the country-level territorial
heterogeneity is fixed (Lessmann and Seidel 2017).

@@

\hypertarget{outline-discussion}{%
\subsubsection{\texorpdfstring{\textbf{Outline
discussion}}{Outline discussion}}\label{outline-discussion}}

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-feldman2014}{}}%
Feldman, Maryann P. 2014. {``The Character of Innovative Places:
Entrepreneurial Strategy, Economic Development, and Prosperity.''}
\emph{Small Business Economics} 43 (1): 9--20.
\url{https://doi.org/10.1007/s11187-014-9574-4}.

\leavevmode\vadjust pre{\hypertarget{ref-gennaioli2014}{}}%
Gennaioli, Nicola, Rafael La Porta, Florencio Lopez De Silanes, and
Andrei Shleifer. 2014. {``Growth in Regions.''} \emph{Journal of
Economic Growth} 19 (3): 259--309.
\url{https://doi.org/10.1007/s10887-014-9105-9}.

\leavevmode\vadjust pre{\hypertarget{ref-lessmann2017}{}}%
Lessmann, Christian, and André Seidel. 2017. {``Regional Inequality,
Convergence, and Its Determinants -- a View from Outer Space.''}
\emph{European Economic Review} 92 (February): 110--32.
\url{https://doi.org/10.1016/j.euroecorev.2016.11.009}.

\leavevmode\vadjust pre{\hypertarget{ref-nguyen2022}{}}%
Nguyen, Thanh Cong. 2022. {``The Effects of Financial Crisis on Income
Inequality.''} \emph{Development Policy Review} 40 (6): e12600.
\url{https://doi.org/10.1111/dpr.12600}.

\end{CSLReferences}
